 This ---
title: "2019_peak-correlation"
author: "Paul Hook"
date: "4/10/2019"
output: html_document
---

First I need to read in the output from running featureCounts on the server side. The "counts" dataframe has a large first comment which is not necessary, so we will skip the first line. The "counts" data frame is large and contains annotation information (1:6) and count information (7:30). These will be split. Sample names are unweildy so they need to be rename in the most automated way possible. End up with final dataframes including "anno", "counts_named", "peak_summary."

```{r load_libraries}
library(here)
library(tidyverse)
library(ggdendro)
library(ggbiplot)
library(patchwork)
library(cqn)
library(sva)
library(formattable)
```

```{r read_in_data}
# Loading data
count.path <- here::here("data","2019_peak-counts.txt")
summary.path <- here::here("data","2019_peak-counts.txt.summary")

count.df <- read.delim(file = count.path, header = T, skip = 1)
summary.df <- read.delim(file = summary.path, header = T)

#Split count dataframe
anno <- count.df %>% dplyr::select(Geneid, Chr, Start, End, Strand, Length)
counts <- count.df %>% dplyr::select(-Geneid, -Chr, -Start, -End, -Strand, -Length)

#Renaming columns of count
samples.bam <- sapply(strsplit(names(counts), "sorted_"), "[", 2)
samples <- sapply(strsplit(samples.bam, ".bam"), "[", 1)
counts_named <- counts
names(counts_named) <- samples

#Renaming columns of summary
peak_summary <- summary.df
names(peak_summary) <- c("Status", samples)
```

Creating a meta data dataframe for all of the samples. This information might be particularly important for trying to determine batch or other covariate effects seen in the data
```{r Creating a pheno_df}
pheno.dir <- here::here("data","peak-annot.txt")
pheno.df <- read.delim(file = pheno.dir, header = T) %>%
  dplyr::arrange(-desc(file.name)) %>%
  mutate(color = case_when(cell.cat == "Glia" ~ "#4DAF4A",
        cell.cat == "Excitatory neurons" ~ "#377EB8",
        cell.cat == 'Inhibitory neurons' ~ "#984EA3",
        cell.cat == 'Dopaminergic neurons' ~ "#E41A1C",
        cell.cat == 'Retina cells' ~ "#FF7F00",
        cell.cat == 'T-cells' ~ "#F781BF"))

```

I first wanted to look at the "raw data" density distributions so I could get a feel as to how to this data looks before normalization. First I copy the "counts_named" object so I do not accidentally modify the original data.
```{r}
Counts <- counts_named
hist(t(log2(Counts+1)), main = "Histogram of log2 transformed counts")

#Changing row names
names(Counts) <- pheno.df$cell.population

counts.log2 <- log2(Counts+1)

log2.cors <- cor(counts.log2, method = "pearson")	
log2.dd <- as.dist(1-abs(log2.cors))
log2.hc <- hclust(log2.dd, method = "ward.D2")
log2.dendro <- ggdendrogram(log2.hc,rotate = T) +
  theme(axis.text.x = element_blank()) + geom_text(size=10)

# Using hierarchial clustering in this way was suggested by http://research.stowers.org/mcm/efg/R/Visualization/cor-cluster/index.htm

# https://cran.r-project.org/web/packages/ggdendro/vignettes/ggdendro.html
# https://stackoverflow.com/questions/8045538/labelling-ggdendro-leaves-in-multiple-colors

# Making dendrogram
dend <- as.dendrogram(log2.hc)
dend_data <- dendro_data(dend,type="rectangle")

labs <- label(dend_data)
labs <- labs %>%
  dplyr::mutate(type = if_else(grepl("microglia|astrocytes|oligo|neun",label,ignore.case = T),"Glia",
                               if_else(grepl("excitatory",label,ignore.case = T),"Excitatory neurons",
                                       if_else(grepl("inhibitory",label,ignore.case = T),"Inhibitory neurons",
                                              if_else(grepl("da",label,ignore.case = T),"Dopaminergic neurons",
                                                      if_else(grepl("cone|rod",label,ignore.case = T),"Retina cells",
                                                              if_else(grepl("T-cells",label,ignore.case = T),"T-cells","NA")))))))

col <- c(Glia="#4DAF4A",
         `Excitatory neurons`="#377EB8",
         `Inhibitory neurons`="#984EA3",
         `Dopaminergic neurons`="#E41A1C",
         `Retina cells`="#FF7F00",
         `T-cells`="#F781BF")

log2.dendro <- ggplot(segment(dend_data)) +
  geom_segment(aes(x=x, y=y, xend=xend, yend=yend),
               size=0.75,lineend = "round",linejoin = "round") +
  geom_text(data=label(dend_data),
               aes(label=label,
                   x=x, y=0,
                   colour=labs$type,
                   angle=45,hjust=1,vjust=1,
                   family="Helvetica",fontface="bold"),
            nudge_y = -0.02) +
  theme_dendro() +
  theme(legend.position="none") +
  ylim(c(-1,1.25)) + xlim(c(-1,26)) +
  scale_color_manual(name = "", values = col)


pca.log2 <- prcomp(t(counts.log2), scale. = T)
log2.out <- as.data.frame(pca.log2$x)
summary(pca.log2)

# Both are ordered by the file name so we can just column bind
final.log2.out <- cbind(log2.out,pheno.df)

p.log2.type <-ggplot(final.log2.out, aes(x=PC1,y=PC2,color=type))
p.log2.type <- p.log2.type+geom_point(size=2) + theme_classic() +
  ggtitle("Type of ATAC-seq") +
  xlab("PC1 (70.29% variance explained)") +
  ylab("PC2 (6.66% variance explained)") +
  theme(axis.title = element_text(size=8),
        axis.text = element_text(size=6),
        legend.text = element_text(size=8),
        plot.title = element_text(hjust = 0.5, face = "bold")) +
  theme(legend.key.size = unit(0.5,"line"),legend.title = element_blank())

p.log2.exp <-ggplot(final.log2.out, aes(x=PC1,y=PC2,color=experiment))
p.log2.exp <- p.log2.exp+geom_point(size=2) + theme_classic() +
  ggtitle("Experiment") + 
  xlab("PC1 (70.29% variance explained)") +
  ylab("PC2 (6.66% variance explained)") +
  theme(axis.title = element_text(size=8),
        axis.text = element_text(size=6),
        axis.title.y = element_blank(),
        legend.text = element_text(size=8),
        plot.title = element_text(hjust = 0.5, face="bold")) +
  theme(legend.key.size = unit(0.5,"line"), legend.title = element_blank())

p.log2.cells <-ggplot(final.log2.out, aes(x=PC1,y=PC2,color=cell.cat))
p.log2.cells <- p.log2.cells+geom_point(size=2) + theme_classic() +
  ggtitle("Cell population") +
  xlab("PC1 (70.29% variance explained)") +
  ylab("PC2 (6.66% variance explained)") +
  theme(axis.title = element_text(size=8),
        axis.text = element_text(size=6),
        legend.text = element_text(size=8),
        axis.title.y = element_blank(),
        plot.title = element_text(hjust = 0.5, face = "bold")) +
  theme(legend.key.size = unit(0.5,"line"), legend.title = element_blank()) +
  scale_color_manual(name = "", values = col)

log2.summary <- log2.dendro / (p.log2.type + p.log2.exp + p.log2.cells)

```

# According to Sarah. there are a few steps in order to set up the data for CQN normalization. Step one is to calculate the sizeFactors which can be done from the "peak_summary" and this will be stored as "sizeFactor."
```{r getting setup for CQN}
# Calculating sizeFactor
sizeFactor <- colSums(peak_summary[,-1])
sizeFactor

# Make Covariate file for CQN including GC content.
# Loading GC-content file
gc_dir <- here::here("data","2019_gc.bed")
gc_content <- read.delim(gc_dir, header = TRUE)

#updating anno file
anno$GC <- gc_content$X6_pct_gc
anno$Length <- gc_content$X13_seq_len

#naming covariates
Covariates <- data.frame(GeneID=anno$Geneid, length=anno$Length, gccontent=anno$GC)

#setting up other CQN things
sizes <- t(sizeFactor)

#Performing quantile normalization with covariates using CQN
normalizing <- cqn(counts = Counts, lengths = Covariates$length, x = Covariates$gccontent, sizeFactors = sizes, verbose = T)

#Modifying the counts
normalized.counts <- normalizing$y + normalizing$offset
hist(normalized.counts, main = "CQN Normalized counts histogram")

#Performing ComBat in order to remove "batch" data
mod0 = model.matrix(~1,data=pheno.df)
combat_data_type <- ComBat(dat=normalized.counts, batch = pheno.df$type, mod=mod0, par.prior = T, prior.plots = F)

# Plotting combined count data
hist(combat_data_type, main = "CQN/ComBat normalaized counts histogram")

#Performing correlation to look at how correlated the samples are
par(mfrow=c(1,1))
combat.cors <- cor(combat_data_type, method = "pearson")	
dd <- as.dist(1-abs(combat.cors))
combat.hc <- hclust(dd, method = "ward.D2")
combat.dendro <- ggdendrogram(combat.hc) + theme(axis.text.y = element_blank())

dend <- as.dendrogram(combat.hc)
dend_data <- dendro_data(dend,type="rectangle")

labs <- label(dend_data)
labs <- labs %>%
  dplyr::mutate(type = if_else(grepl("microglia|astrocytes|oligo|neun",label,ignore.case = T),"Glia",
                               if_else(grepl("excitatory",label,ignore.case = T),"Excitatory neurons",
                                       if_else(grepl("inhibitory",label,ignore.case = T),"Inhibitory neurons",
                                              if_else(grepl("da",label,ignore.case = T),"Dopaminergic neurons",
                                                      if_else(grepl("cone|rod",label,ignore.case = T),"Retina cells",
                                                              if_else(grepl("T-cells",label,ignore.case = T),"T-cells","NA")))))))

col <- c(Glia="#4DAF4A",
         `Excitatory neurons`="#377EB8",
         `Inhibitory neurons`="#984EA3",
         `Dopaminergic neurons`="#E41A1C",
         `Retina cells`="#FF7F00",
         `T-cells`="#F781BF")

combat.dendro <- ggplot(segment(dend_data)) +
  geom_segment(aes(x=x, y=y, xend=xend, yend=yend),
               size=0.75,lineend = "round",linejoin = "round") +
  geom_text(data=label(dend_data),
               aes(label=label,
                   x=x, y=0,
                   colour=labs$type,
                   angle=90,hjust=1.05,vjust=0.5,
                   family="Helvetica",fontface="bold"),
            nudge_y = -0.02,size=2.5) +
  theme_dendro() +
  theme(legend.position="none",
        text = element_text(family = "Helvetica")) +
  ylim(c(-1,2)) + xlim(c(-1,26)) +
  scale_color_manual(name = "", values = col) +
  theme(plot.margin = margin(0, -2, 0,-2, "cm"))
  NULL

#Performing PCA 
pca.combat <- prcomp(t(combat_data_type), scale. = TRUE)
combat.out <- as.data.frame(pca.combat$x)

final.combat.out <- cbind(combat.out,pheno.df)

#Getting the summary data from the PCA, especially performed the get the proportion of variance
summary(pca.combat)

#Plotting PC1 vs. PC2, colored by type of experiment
#Plotting
p.combat.type <-ggplot(final.combat.out, aes(x=PC1,y=PC2,color=type))
p.combat.type<- p.combat.type+geom_point(size=2) + theme_classic() +
  #ggtitle("Type of ATAC-seq") +
  xlab("PC1 (19.73% variance explained)") +
  ylab("PC2 (11.8% variance explained)") +
  theme(axis.title = element_text(size=8),
        axis.text = element_text(size=6),
        legend.text = element_text(size=8),
        plot.title = element_text(hjust = 0.5, face = "bold")) +
  theme(legend.key.size = unit(0.5,"line"),legend.title = element_blank())


p.combat.exp <-ggplot(final.combat.out, aes(x=PC1,y=PC2,color=experiment))
p.combat.exp<- p.combat.exp+geom_point(size=2) + theme_classic() +
  #ggtitle("Experiment") + 
  xlab("PC1 (19.73% variance explained)") +
  #ylab("PC2 (11.8% variance explained)") +
  theme(axis.title = element_text(size=8),
        axis.title.y = element_blank(),
        axis.text = element_text(size=6),
        legend.text = element_text(size=8),
        plot.title = element_text(hjust = 0.5, face = "bold")) +
  theme(legend.key.size = unit(0.5,"line"),legend.title = element_blank())


p.combat.cells <-ggplot(final.combat.out, aes(x=PC1,y=PC2,color=cell.cat))
p.combat.cells<- p.combat.cells+geom_point(size=2) + theme_classic() +
  #ggtitle("Cell population") +
  xlab("PC1 (19.73% variance explained)") +
  #ylab("PC2 (11.8% variance explained)") +
  theme(axis.title = element_text(size=8),
        axis.title.y = element_blank(),
        axis.text = element_text(size=6),
        legend.text = element_text(size=8),
        plot.title = element_text(hjust = 0.5, face = "bold")) +
  theme(legend.key.size = unit(0.5,"line"),legend.title = element_blank()) +
  scale_color_manual(name = "", values = col)
  
combat.summary <- p.combat.type + p.combat.exp + p.combat.cells

combat.dendro + p.combat.cells

ggsave(combat.dendro,
       filename = here::here("Paper","results_1","2019-combat-dendrogram.png"),
       width = 10, height=5,dpi = "print")

ggsave(p.combat.cells,
       filename = here::here("Paper","results_1","2019-combat-PCA.png"),
       width = 6, height=5,dpi = "print")
```

```{r}
scree.combat <- ggscreeplot(pca.combat, type = "pev") +
  ggtitle('Screeplot') + theme_classic() +
  theme(axis.title = element_text(size=8),
        plot.title = element_text(hjust = 0.5, face = "bold"))
```

```{r}
library(tsne)

nComponents <- 6
seed <- 1834
set.seed(seed) #setting seed

combat.tsne<-tsne(pca.combat$x[,1:nComponents],perplexity=5,max_iter=10000,whiten = T)

pheno.df$tsne1 <- combat.tsne[,1]
pheno.df$tsne2 <- combat.tsne[,2]

col <- c(Glia="#4DAF4A",
         `Excitatory neurons`="#377EB8",
         `Inhibitory neurons`="#984EA3",
         `Dopaminergic neurons`="#E41A1C",
         `Retina cells`="#FF7F00",
         `T-cells`="#F781BF")

tsne.cat <-ggplot(pheno.df, aes(x=tsne1,y=tsne2,color=cell.cat,label=cell.population)) +
  geom_point(size=3) + theme_classic() +
  xlab("t-SNE 1") + ylab("t-SNE 2") +
  theme(axis.text = element_blank(), 
        axis.ticks = element_blank(),
        legend.position = "right",
        axis.title = element_text(face="bold",family = "Helvetica"),
        legend.text=element_text(size=8),
        legend.text.align = 0,
        legend.key.height = unit(0.5,'cm'),
        legend.title = element_text(size=10),
        legend.title.align = 0) +
  scale_color_manual(name="Broad Cell Population", values = col) +
  geom_text_repel(data=subset(pheno.df, cell.population == "Inhibitory*" | cell.population == "Inhibitory MSN*"),
            aes(tsne1,tsne2,label=cell.population),family="Helvetica",size=2.5,hjust=0,vjust=0,seed = 24)

final.tsne.cat <- tsne.cat + theme(plot.margin = margin(0.5, 3, 0.5, 3, "cm"))

fig1a <- combat.dendro + final.tsne.cat

#pdf(file = "Figure1A_dendro-tsne.pdf",height = 4, width = 8)
#fig1a
#dev.off()

pdf(file = "Figure1A_dendro-tsne_V2.pdf",height = 4, width = 8)
fig1a
dev.off()
```

# Make figure S1
```{r}
pca.summaries <- (p.log2.type + p.log2.exp + p.log2.cells) / (p.combat.type + p.combat.exp + p.combat.cells)

pdf(file = "Figure.S1_pca.pdf",height =4,width =10)
pca.summaries
dev.off()
```

