---
title: "all-pairwise"
author: "Paul Hook"
date: "10/24/2019"
output: html_document
---

Loading the required libraries
```{r, libraries}
library(dplyr)
library(tibble)
library(stringr)
library(readr)
library(dendextend)
library(gplots)
library(RColorBrewer)
library(pheatmap)
```

Reading in the Jaccard index matrices as well as data that will be used to annotate the rows and columns.
```{r load-data, include=FALSE}
# Loading in the jaccard data
x <- read.table("../final.data/all.matrix")

# Loading in the Epigenome Roadmap Atlas cell population names
eid <- readr::read_tsv("../EIDlegend_mod.txt")

# Loading in the ATAC-seq metadata and modifying names
cell_annot <- readr::read_tsv("../cell-annot.txt") %>%
  dplyr::mutate(filename=str_remove(filename,"summits_filter2_")) %>%
  dplyr::mutate(filename=str_remove(filename,"_summits")) %>%
  dplyr::mutate(filename=str_replace_all(filename,"-","."))
```

Now we need to modify the jaccard matrix to just keep all ATAC-DNAse comparisons. ATAC profiles will be columns and DNAse will be rows so we can make a nice looking heatmap
```{r modify-matrix}
# Modifying the count matrix so columns are ATAC-seq populations and rows are DNAse
all.names <- x %>%
  # Only keep columns that include data for ATAC-seq populations
  dplyr::select(-ends_with("DNase.")) %>%
  # Move the rownames into a column
  tibble::rownames_to_column('eid') %>%
  # Only keep rows for DNAse
  dplyr::filter(str_detect(eid,"DNase.")) %>%
  # Remove "-DNase." from names
  dplyr::mutate(eid=str_remove(eid,"-DNase.")) %>%
  # Join matrix to Roadmap names so the names are actually used
  dplyr::left_join(.,eid,by="eid") %>%
  # Only keep things you want
  dplyr::select(Name,everything(),-eid,-roadmap_name,-Category,-Type) %>%
  {.}

# Make rownames the Roadmap names
row.names(all.names) <- all.names$Name
all.names <- all.names[,-1]

# Now annotate the ATAC-seq samples with the names used in the paper
# First transpose the matrix
flip <- tibble::as_tibble(t(all.names),rownames = "filename") %>%
  # Join it to the cell annoation
  dplyr::left_join(.,cell_annot,by="filename") %>%
  # Select columns we want to keep
  dplyr::select(Name,everything(),-filename,-cells) %>%
  # Change column to rownames and transpose back
  tibble::column_to_rownames('Name') %>% t()
```

Creating eventual heatmap annotation
```{r heatmap-annotation}
# Create column annotation for the eventual heatmap
# In the end, we need to replicate what happens when we scale the date in the headmap
# Here we are flipping the matrix and calculating the Z-scores within an ATAC-seq sample
tmp <- t(flip)
scale_rows = function(x){
    m = apply(x, 1, mean, na.rm = T)
    s = apply(x, 1, sd, na.rm = T)
    return((x - m) / s)
}
tmp2 <- scale_rows(tmp)

# Now we need to cluster the data like the heatmap will
# Performing heirarchial clustering
my_hclust_eid<- hclust(dist(t(tmp2)), method = "ward.D")
# Making a dataframe so clustering results can be used as annotation
my_eid_col <- dendextend::cutree(tree=as.dendrogram(my_hclust_eid),
                                 k=3,order_clusters_as_data = FALSE)
my_eid_col <- as.data.frame(my_eid_col) %>%
  tibble::rownames_to_column('Name')

# Merge the heirarchial clustering with the Roadmap names
test <- merge(my_eid_col,eid,by='Name',sort = FALSE) %>% 
  dplyr::select(Name,"Roadmap category"=Category) %>% 
  tibble::column_to_rownames("Name") 

# Creating row annotation
# Doing the same thing above but with ATAC-seq datasets
# Performing heirarchial clustering
my_hclust_cells<- hclust(dist(tmp2), method = "ward.D")
# Making a dataframe so clustering results can be used as annotation
my_cells_col <- dendextend::cutree(tree=as.dendrogram(my_hclust_cells),
                                   k=3,order_clusters_as_data = FALSE)
my_cells_col <- as.data.frame(my_cells_col) %>%
  tibble::rownames_to_column('Name')

# Merge the heirarchial clustering with the ATAC cell annotations
cells.dat <- merge(my_cells_col, cell_annot, by='Name',sort = FALSE) %>% 
  dplyr::select(Name,"type"=cells) %>% 
  tibble::column_to_rownames("Name") 
```

Now time to make the heatmap
```{r heatmap}
# Setting the colors for the annotations
ann_colors = list(
    type = c(Glia="#4DAF4A",
              `Excitatory neurons`="#377EB8",
              `Inhibitory neurons`="#984EA3",
              `Dopaminergic neurons`="#E41A1C",
              `Retina cells`="#FF7F00",
              `T-cells`="#F781BF"),
    `Roadmap category` = RColorBrewer::brewer.pal(11,"Set3")
)
# Setting the factor levels for Roadmap categories
names(ann_colors$`Roadmap category`) <- levels(factor(test$`Roadmap category`))

# Setting the palette for the heatmap
pal <- colorRampPalette(rev(brewer.pal(11,"RdBu")))

# Making the heatmap
pheno.map <- pheatmap(flip,
         clustering_method = "ward.D",
         cluster_rows = T, cluster_cols = T,
         color = pal(100),
         scale = "column",
         cellwidth = 11,
         cellheight = 10,
         main="All Imputed Roadmap DNase",
         number_color = "black",
         fontsize_number = 10,
         fontface="bold",
         show_colnames = TRUE,
         border_color = "NA",
         annotation_col = cells.dat,
         annotation_names_col = FALSE, annotation_names_row = FALSE,
         annotation_row = test,
         fontsize_col = 8,
         fontsize_row = 8,
         annotation_colors = ann_colors,
         angle_col = "45",
         fontfamily="Helvetica")

# Printing the heatmap
pdf(file = "../final.results/all-heatmap.pdf",width = 12, height = 20)
pheno.map
dev.off()
```

Making a heat map that will better fit on a published figure
```{r}
new.map <- pheatmap(flip,
         clustering_method = "ward.D",
         cluster_rows = T, cluster_cols = T,
         color = pal(100),
         scale = "column",fontsize = 8,
         cellwidth = 8,
         cellheight = 2,
         main="Roadmap promoters",
         number_color = "black",
         fontsize_number = 10,
         fontface="bold",
         show_colnames = TRUE,
         border_color = "NA",
         annotation_col = cells.dat,
         annotation_names_col = FALSE, annotation_names_row = FALSE,
         annotation_row = test,
         fontsize_col = 6,
         fontsize_row = 2,
         annotation_colors = ann_colors,
         angle_col = "45",
         fontfamily="Helvetica")
new.map

pdf(file = "../final.results/publication-promoter-heatmap.pdf",width = 10, height = 6)
new.map
dev.off()
```

# Need to modify the figure above to be smaller for publication
```{r}
new.map <- pheatmap(flip,
         clustering_method = "ward.D",
         cluster_rows = T, cluster_cols = T,
         color = pal(100),
         scale = "column",fontsize = 8,
         cellwidth = 10,
         cellheight = 2,
         main="All Imputed Roadmap DNase",
         number_color = "black",
         fontsize_number = 10,
         fontface="bold",
         show_colnames = TRUE,
         border_color = "NA",
         annotation_col = cells.dat,
         annotation_names_col = FALSE, annotation_names_row = FALSE,
         annotation_row = test,
         fontsize_col = 8,
         fontsize_row = 2,
         annotation_colors = ann_colors,
         angle_col = "45",
         fontfamily="Helvetica")
new.map

pdf(file = "../final.results/publication-all-heatmap.pdf",width = 10, height = 6)
new.map
dev.off()
```