---
title: "atac-2-atac"
author: "Paul Hook"
date: "12/10/2019"
output: html_document
---

Load the required libraries
```{r, libraries}
library(dplyr)
library(tibble)
library(stringr)
library(readr)
library(dendextend)
library(gplots)
library(RColorBrewer)
library(pheatmap)
```

Reading in the Jaccard index matrices as well as data that will be used to annotate the rows and columns. For these comparisons we could use whichever dataset we wanted because all ATAC-ATAC comparisons were thrown out in the other analyses (all, promoter, enhancer, dyadic)
```{r setup, include=FALSE}
# Loading in the jaccard data
x <- read.table("../final.data/enh.matrix")

# Loading in the ATAC-seq metadata and modifying names
cell_annot <- readr::read_tsv("../cell-annot.txt") %>%
  dplyr::mutate(filename=str_remove(filename,"summits_filter2_")) %>%
  dplyr::mutate(filename=str_remove(filename,"_summits")) %>%
  dplyr::mutate(filename2=str_replace_all(filename,"-","."))
```

Now we need to modify the jaccard matrix to just keep all ATAC-ATAC comparisons
```{r}
# Modifying the count matrix so columns are ATAC-seq populations and rows are DNAse
atac <- x %>%
  # Only keep columns that include data for ATAC-seq populations
  dplyr::select(-starts_with("regions")) %>%
  # Move the rownames into a column
  tibble::rownames_to_column('filename') %>%
  # Only keep rows for ATAC
  dplyr::filter(str_detect(filename,"regions",negate = TRUE)) %>%
  # Join matrix to ATAC names so the names are actually used
  dplyr::left_join(.,cell_annot,by="filename") %>%
  # Only keep things you want
  dplyr::select(Name,everything(),-cells,-filename,-filename2) %>%
  # Change column to rownames
  tibble::column_to_rownames('Name') %>%
  {.}

# Do the same annotation so both sides of the matrix have the cell names
atac.mod <- tibble::as_tibble(t(atac),rownames = "filename2") %>%
  dplyr::left_join(.,cell_annot,by="filename2") %>%
  dplyr::select(Name,everything(),-cells,-filename,-filename2) %>%
  tibble::column_to_rownames('Name') %>%
  {.}
```

Creating eventual heatmap annotation for the raw Jaccard statistics
```{r heatmap-annotation}
# Create annotation for the eventual heatmap
# This annotation should be the same for rows and columns
# In the end, we need to replicate what happens when the heatmap program clusters the data
# Creating a temporary matrix
tmp <- atac.mod

# Now we need to cluster the data like the heatmap will
# Performing heirarchial clustering
atac.col <- hclust(dist(atac.mod), method = "ward.D")
my_atac_col <- dendextend::cutree(tree=as.dendrogram(atac.col),
                                  k=3,order_clusters_as_data = FALSE)
my_atac_col <- as.data.frame(my_atac_col) %>%
  tibble::rownames_to_column('Name')

# Merge the heirarchial clustering with the ATAC names
test <- merge(my_atac_col,cell_annot,by='Name',sort = FALSE) %>% 
  dplyr::select(Name,"Type"=cells) %>% 
  tibble::column_to_rownames("Name") 
```

Now time to make the heatmap for raw Jaccard statistics
```{r raw-jaccard}
# Setting the colors for the annotations
ann_colors = list(
    Type = c(Glia="#4DAF4A",
              `Excitatory neurons`="#377EB8",
              `Inhibitory neurons`="#984EA3",
              `Dopaminergic neurons`="#E41A1C",
              `Retina cells`="#FF7F00",
              `T-cells`="#F781BF")
)

# Setting the palette for the heatmap
pal.1 <- colorRampPalette(brewer.pal(9,"Reds"))

# Making the heatmap
pheno.map <- pheatmap(atac.mod,
         clustering_method = "ward.D",
         cluster_rows = T, cluster_cols = T,
         color = pal.1(100),
         cellwidth = 6,
         cellheight = 6,
         fontsize = 6,
         main="Jaccard",
         angle_col = "90",
         number_color = "black",
         fontsize_number = 10,
         fontface="bold",
         show_colnames = TRUE,
         border_color = "NA",
         fontsize_col = 6,
         fontsize_row = 6,
         annotation_colors = ann_colors,
         annotation_row = test,
         annotation_col = test,
         annotation_names_row = FALSE,annotation_names_col = FALSE)

# Printing the heatmap
pdf(file = "../final.results/atac-jaccard.pdf",width = 6, height = 5)
pheno.map
dev.off()
```

Now we want to make a heatmap for "scaled" jaccard statistics
```{r}
# Create column annotation for the eventual heatmap
# In the end, we need to replicate what happens when we scale the data in the heatmap
# This annotation should be the same for rows and columns
# Creating a scaled matrix
scale.atac <- scale(atac.mod)

# Now we need to cluster the data like the heatmap will
# Performing heirarchial clustering
atac.col <- hclust(dist(scale.atac), method = "ward.D")
# Making a dataframe so clustering results can be used as annotation
my_atac_col <- dendextend::cutree(tree=as.dendrogram(atac.col),
                                  k=3,order_clusters_as_data = FALSE)
my_atac_col <- as.data.frame(my_atac_col) %>%
  tibble::rownames_to_column('Name')

# Merge the heirarchial clustering with the ATAC names
test <- merge(my_atac_col,cell_annot,by='Name',sort = FALSE) %>% 
  dplyr::select(Name,"Type"=cells) %>% 
  tibble::column_to_rownames("Name") 
```

Now time to make the heatmap for "scaled" Jaccard statistics
```{r scaled-jaccard}
# Setting the colors for the annotations
ann_colors = list(
    Type = c(Glia="#4DAF4A",
              `Excitatory neurons`="#377EB8",
              `Inhibitory neurons`="#984EA3",
              `Dopaminergic neurons`="#E41A1C",
              `Retina cells`="#FF7F00",
              `T-cells`="#F781BF")
)

# Setting the palette for the heatmap
pal.2 <- colorRampPalette(rev(brewer.pal(11,"RdBu")))

# Making the heatmap
pheno.map <- pheatmap(atac.mod,
         clustering_method = "ward.D",
         cluster_rows = T, cluster_cols = T,
         color = pal.2(100),
         scale = "column",
         cellwidth = 6,
         cellheight = 6,
         fontsize = 6,
         main="Scaled Jaccard",
         angle_col = "90",
         number_color = "black",
         fontsize_number = 10,
         fontface="bold",
         show_colnames = TRUE,
         border_color = "NA",
         fontsize_col = 6,
         fontsize_row = 6,
         annotation_colors = ann_colors,
         annotation_row = test,
         annotation_col = test,
         annotation_names_row = FALSE,annotation_names_col = FALSE)

# Printing heatmap
pdf(file = "../final.results/scale-atac-jaccard.pdf",width = 6, height = 5)
pheno.map
dev.off()

```

Now time to make the heatmap showing the correlation of Jaccard statistics between samples
```{r}
# Calculating the spearman correlation between all profiles
cor.atac <- cor(atac.mod,method = "spearman")

# Now we need to cluster the data like the heatmap will
# Performing heirarchial clustering
atac.col <- hclust(dist(cor.atac), method = "ward.D")
my_atac_col <- dendextend::cutree(tree=as.dendrogram(atac.col),
                                  k=3,order_clusters_as_data = FALSE)
# Making a dataframe so clustering results can be used as annotation
my_atac_col <- as.data.frame(my_atac_col) %>%
  tibble::rownames_to_column('Name')

# Merge the heirarchial clustering with the ATAC names
test <- merge(my_atac_col,cell_annot,by='Name',sort = FALSE) %>% 
  dplyr::select(Name,"Type"=cells) %>% 
  tibble::column_to_rownames("Name") 
```

```{r correlation-heatmap}
# Setting the palette for the heatmap
pal.cor <- colorRampPalette(rev(brewer.pal(11,"RdGy")))

# Making the heatmap
cor.map <- pheatmap(cor.atac,
         clustering_method = "ward.D",
         cluster_rows = T, cluster_cols = T,
         color = pal.2(100),
         cellwidth = 6,
         cellheight = 6,
         fontsize = 6,
         main="Jaccard correlation",
         angle_col = "90",
         number_color = "black",
         fontsize_number = 10,
         fontface="bold",
         show_colnames = TRUE,
         border_color = "NA",
         fontsize_col = 6,
         fontsize_row = 6,
         annotation_colors = ann_colors,
         annotation_row = test,
         annotation_col = test,
         legend_breaks = seq(-1, 1, length.out = 11))

# Print results
pdf(file = "../final.results/correlation-atac-jaccard.pdf",width = 6, height = 5)
cor.map
dev.off()
```
